Retrofitting Effect Handlers onto OCaml

### 2020-11-17 retro-concurrency.tex
edits

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperfom is implemented as a
primitive to avoid executing code on the resumption path.

@@

In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
fiber in the current stack. \textsc{AdminO} performs administrative reductions.
\textsc{CallO} evaluates an OCaml function on the current stack.
\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
on an empty C stack with the current OCaml stack as its tail. \textsc{RetToC}
returns a value to the enclosing C stack. In this case, we have exactly one
fiber on the stack, and this was created in the rule \textsc{Callback}, whose
handler has identity return case alone and the environment is empty.
\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
body of the return case.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation |k''|.
Observe that the captured continuation |k''| includes the current handler.
Intuitively, the handler wraps around captured continuation. This gives
Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
captured continuation |k''| in the process. Recall that we do not capture C
frames as part of a continuation. To this end, \textsc{EffUnHn} models
unhandled effect. If the effect bubbles up to the top fiber, which we know does
not have an effect case (see \textsc{Callback} rule), we raise
\textsc{Unhandled} exception at the point where the corresponding effect was
performed. This is achieved by appending the captured continuation to the front
of the current continuation.

>>>>

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns an |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect an |'a| value for
resumption and return a |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperform is implemented as a
primitive to avoid executing code on the resumption path.

@@

In OCaml (Figure~\ref{sem:ostep}), reductions always occur on the top-most
fiber in the current stack. \textsc{AdminO} performs administrative reductions.
\textsc{CallO} evaluates an OCaml function on the current stack.
\textsc{ExtCall} captures the behaviour of external calls, which are evaluated
on an empty C stack with the current OCaml stack as its tail. \textsc{RetToC}
returns a value to the enclosing C stack. In this case, we have exactly one
fiber on the stack, and this was created in the rule \textsc{Callback}, whose
handler has identity return case alone and the environment is empty.
\textsc{RetFib} returns the value from a fiber to the previous one, evaluating the
body of the return case.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation |k''|.
Observe that the captured continuation |k''| includes the current handler.
Intuitively, the handler wraps around captured continuation. This gives
Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
captured continuation |k''| in the process. Recall that we do not capture C
frames as part of a continuation. To this end, \textsc{EffUnHn} models
unhandled effect. If the effect bubbles up to the top fiber --- which we know does
not have an effect case (see \textsc{Callback} rule) --- we raise
\textsc{Unhandled} exception at the point where the corresponding effect was
performed. This is achieved by appending the captured continuation to the front
of the current continuation.
